---
title: Using 2Delta Puts to Hedge Short Strangles
description: Analyzing how long 2delta puts would impact short strangle returns on SPY
author: Ned H
date: 2024-12-15
output:
  github_document:
    toc: true
    toc_depth: 3
    html_preview: false
    df_print: kable
file_path: apps\spy-strangles-with-put-hedge\main.Rmd
---

```{r include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    warning = FALSE,
    message = FALSE,
    fig.height = 6,
    fig.width = 12,
    fig.align = "center",
    fig.path = "dist/images/"
)
```

## Setup

### Dependencies

```{r}
library(tidyverse)

source("packages/utils/aesthetics.R")
```

### Query base strangle data from big query

```{r}
bq_tbl <- read_csv(
    file = "apps/spy-strangles-with-put-hedge/data/bq-strangles-data.csv",
    show_col_types = FALSE
)

bq_tbl |>
    head(3)
```

### Calculate PnLs

```{r}
pnls_tbl <- bq_tbl |>
    mutate(
        strangle_dollar_pnl = strangle_price_exit - strangle_price_entry,
        short_strangle_dollar_pnl = strangle_dollar_pnl * -1,
        tail_put_dollar_pnl = tail_put_price_exit - tail_put_price_entry,
        position_dollar_pnl = short_strangle_dollar_pnl + tail_put_dollar_pnl
    )

pnls_tbl |>
    head(3)
```

## Exploratory Data Analysis

### Tail Put Validation

Series of tests to confirm that the tail put makes sense.

```{r}
# Extracting the strike from the option's id
tail_put_tbl <- pnls_tbl |>
    mutate(tail_put_strike = as.numeric(substr(tail_put_id, nchar(tail_put_id) - 7, nchar(tail_put_id))) / 1000)
```

When stock exit prices goes lower than put strike, all tail_put_dollar_pnl is greater than 0.

```{r}
tail_put_tbl |>
    filter(tail_put_strike > stock_price_exit) |>
    select(tail_put_dollar_pnl) |>
    summary()
```

There's only been 8 instances since 2019 when the stock_price_exit passed the put strike.

```{r}
tail_put_tbl |>
    count(tail_put_strike > stock_price_exit)
```

The tail put hedge has only made money 11 times.

```{r}
tail_put_tbl |>
    count(tail_put_dollar_pnl > 0)
```

The 3 times that tail put hedge pnl was positive that wasn't when the underlying was less than or equal to the put's strike is when it was very close to it.

```{r}
tail_put_tbl |>
    filter(tail_put_strike <= stock_price_exit &
        tail_put_dollar_pnl >= 0)
```

For example, the mean percentage of the difference between the put's strike and the stock's exit price with the put's strike is ~0.0008.

```{r}
tail_put_tbl |>
    filter(tail_put_strike <= stock_price_exit &
        tail_put_dollar_pnl >= 0) |>
    mutate(diff = stock_price_exit - tail_put_strike) |>
    select(tail_put_dollar_pnl, diff, tail_put_strike) |>
    mutate(pct_diff = diff / tail_put_strike) |>
    summary()
```

### Short Strangles Delta

TODO

### Tail Puts Delta

```{r}
put_delta_summaries <- pnls_tbl |>
    select(date_entry, tail_put_delta_entry) |>
    reframe(
        min_date = min(date_entry),
        max_date = max(date_entry),
        med_delta = median(tail_put_delta_entry),
        mean_delta = mean(tail_put_delta_entry),
        min_delta = min(tail_put_delta_entry),
        max_delta = max(tail_put_delta_entry),
        sd_delta = sd(tail_put_delta_entry),
        n = n()
    ) |>
    mutate(across(contains("delta"), function(x) round(x, 3)))

tail_put_deltas_plot <- pnls_tbl |>
    select(date_entry, tail_put_delta_entry) |>
    ggplot(aes(tail_put_delta_entry, fill = "name")) +
    geom_histogram(bins = 100) +
    labs(
        title = "Tail Put Deltas at Entry",
        subtitle = sprintf(
            "SPY 7dte put deltas from %s to %s (n=%s, µ=%s, min=%s, max=%s, sd=%s)",
            put_delta_summaries$min_date,
            put_delta_summaries$max_date,
            put_delta_summaries$n,
            put_delta_summaries$mean_delta,
            put_delta_summaries$min_delta,
            put_delta_summaries$max_delta,
            put_delta_summaries$sd_delta
        ),
        x = "Put Delta",
        y = "Frequency"
    )

tail_put_deltas_plot +
    scale_fill_manual(values = viridis::viridis(1, option = "E")) +
    theme(legend.position = "none")
```

```{r include=FALSE, eval=FALSE}
styled_tail_put_deltas_plot <- tail_put_deltas_plot +
    scale_fill_manual(values = nedhmn_palette) +
    nedhmn_theme()

ggsave(
    filename = "apps/spy-strangles-with-put-hedge/public/tail-put-delta-distribution.png",
    plot = styled_tail_put_deltas_plot,
    height = 4,
    width = 8
)
```

### Tail Put Entry Prices

```{r}
put_price_summaries <- pnls_tbl |>
    select(date_entry, tail_put_price_entry) |>
    reframe(
        min_date = min(date_entry),
        max_date = max(date_entry),
        med = median(tail_put_price_entry),
        mean = mean(tail_put_price_entry),
        min = min(tail_put_price_entry),
        max = max(tail_put_price_entry),
        sd = sd(tail_put_price_entry),
        n = n()
    ) |>
    mutate(across(med:sd, function(x) scales::dollar(round(x, 3))))

put_prices_plot <- pnls_tbl |>
    select(date_entry, tail_put_price_entry) |>
    ggplot(aes(tail_put_price_entry, fill = "name")) +
    geom_histogram(bins = 100) +
    scale_x_continuous(labels = scales::dollar) +
    labs(
        title = "Tail Put Prices at Entry",
        subtitle = sprintf(
            "SPY 7dte put prices from %s to %s (n=%s, µ=%s, min=%s, max=%s, sd=%s)",
            put_price_summaries$min_date,
            put_price_summaries$max_date,
            put_price_summaries$n,
            put_price_summaries$mean,
            put_price_summaries$min,
            put_price_summaries$max,
            put_price_summaries$sd
        ),
        x = "Option Price",
        y = "Frequency"
    )

put_prices_plot +
    scale_fill_manual(values = viridis::viridis(1, option = "E")) +
    theme(legend.position = "none")
```

```{r include=FALSE}
styled_put_prices_plot <- put_prices_plot +
    scale_fill_manual(values = nedhmn_palette) +
    nedhmn_theme()

ggsave(
    filename = "apps/spy-strangles-with-put-hedge/public/put_prices_distribution.png",
    plot = styled_put_prices_plot,
    height = 4,
    width = 8
)
```

### Strangle Entry Prices

```{r}
strangle_price_summaries <- pnls_tbl |>
    select(date_entry, strangle_price_entry) |>
    reframe(
        min_date = min(date_entry),
        max_date = max(date_entry),
        med = median(strangle_price_entry),
        mean = mean(strangle_price_entry),
        min = min(strangle_price_entry),
        max = max(strangle_price_entry),
        sd = sd(strangle_price_entry),
        n = n()
    ) |>
    mutate(across(med:sd, function(x) scales::dollar(round(x, 3))))

strangle_prices_plot <- pnls_tbl |>
    select(date_entry, strangle_price_entry) |>
    ggplot(aes(strangle_price_entry, fill = "name")) +
    geom_histogram(bins = 100) +
    scale_x_continuous(labels = scales::dollar) +
    labs(
        title = "Strangle Prices at Entry",
        subtitle = sprintf(
            "SPY 7dte strangle prices from %s to %s (n=%s, µ=%s, min=%s, max=%s, sd=%s)",
            strangle_price_summaries$min_date,
            strangle_price_summaries$max_date,
            strangle_price_summaries$n,
            strangle_price_summaries$mean,
            strangle_price_summaries$min,
            strangle_price_summaries$max,
            strangle_price_summaries$sd
        ),
        x = "Option Price",
        y = "Frequency"
    )

strangle_prices_plot +
    scale_fill_manual(values = viridis::viridis(1, option = "E")) +
    theme(legend.position = "none")
```

```{r include=FALSE}
styled_strangle_prices_plot <- strangle_prices_plot +
    scale_fill_manual(values = nedhmn_palette) +
    nedhmn_theme()

ggsave(
    filename = "apps/spy-strangles-with-put-hedge/public/strangle_prices_distribution.png",
    plot = styled_strangle_prices_plot,
    height = 4,
    width = 8
)
```

### Short Strangle PnL

```{r}
short_strangle_pnl_summaries <- pnls_tbl |>
    select(date_entry, short_strangle_dollar_pnl) |>
    reframe(
        min_date = min(date_entry),
        max_date = max(date_entry),
        min = min(short_strangle_dollar_pnl),
        mean = mean(short_strangle_dollar_pnl),
        max = max(short_strangle_dollar_pnl),
        sd = sd(short_strangle_dollar_pnl),
        n = n()
    ) |>
    mutate(across(min:max, function(x) scales::dollar(x)))

short_strangle_pnl_plot <- pnls_tbl |>
    ggplot(aes(short_strangle_dollar_pnl * 100, fill = "name")) +
    geom_histogram(bins = 100) +
    scale_x_continuous(labels = scales::dollar) +
    labs(
        title = "SPY Short Strangle Dollar PnL",
        subtitle = sprintf(
            "7DTE 20Delta Short Strangles from %s to %s (n=%s, µ=%s, min=%s, max=%s, sd=%s)",
            short_strangle_pnl_summaries$min_date,
            short_strangle_pnl_summaries$max_date,
            short_strangle_pnl_summaries$n,
            short_strangle_pnl_summaries$mean,
            short_strangle_pnl_summaries$min,
            short_strangle_pnl_summaries$max,
            short_strangle_pnl_summaries$sd
        ),
        x = "PnL",
        y = "Frequency"
    )

short_strangle_pnl_plot +
    scale_fill_manual(values = viridis::viridis(1, option = "E")) +
    theme(legend.position = "none")
```

```{r include=FALSE, eval=FALSE}
styled_short_strangle_pnl_plot <- short_strangle_pnl_plot +
    scale_fill_manual(values = nedhmn_palette) +
    nedhmn_theme()

ggsave(
    filename = "apps/spy-strangles-with-put-hedge/public/short-strangles-distribution.png",
    plot = styled_short_strangle_pnl_plot,
    height = 4,
    width = 8
)
```

### Long Tail Put PnL

```{r}
tail_put_summaries <- pnls_tbl |>
    select(date_entry, tail_put_dollar_pnl) |>
    mutate(tail_put_dollar_pnl = tail_put_dollar_pnl * 100) |>
    reframe(
        min_date = min(date_entry),
        max_date = max(date_entry),
        med = median(tail_put_dollar_pnl),
        mean = mean(tail_put_dollar_pnl),
        min = min(tail_put_dollar_pnl),
        max = max(tail_put_dollar_pnl),
        sd = round(sd(tail_put_dollar_pnl), 2),
        n = n()
    ) |>
    mutate(across(med:max, function(x) scales::dollar(x)))

tail_put_pnl_plot <- pnls_tbl |>
    ggplot(aes(tail_put_dollar_pnl * 100, fill = "name")) +
    geom_histogram(bins = 100) +
    scale_x_continuous(labels = scales::dollar) +
    labs(
        title = "Long SPY Tail Put Dollar PnL",
        subtitle = sprintf(
            "7dte 2delta puts from %s to %s (n=%s, µ=%s, min=%s, max=%s, sd=%s)",
            tail_put_summaries$min_date,
            tail_put_summaries$max_date,
            tail_put_summaries$n,
            tail_put_summaries$mean,
            tail_put_summaries$min,
            tail_put_summaries$max,
            tail_put_summaries$sd
        ),
        x = "PnL",
        y = "Frequency"
    )

tail_put_pnl_plot +
    scale_fill_manual(values = viridis::viridis(1, option = "E")) +
    theme(legend.position = "none")
```

```{r include=FALSE, eval=FALSE}
styled_tail_put_pnl_plot <- tail_put_pnl_plot +
    scale_fill_manual(values = nedhmn_palette) +
    nedhmn_theme()

ggsave(
    filename = "apps/spy-strangles-with-put-hedge/public/long-tail-put-pnl-distribution.png",
    plot = styled_tail_put_pnl_plot,
    height = 4,
    width = 8
)
```

### Put Percent Of Short Strangles

```{r}
pct_of_strangle_summaries <- pnls_tbl |>
    select(date_entry, strangle_price_entry, tail_put_price_entry) |>
    mutate(put_pct_of_strangles = abs(tail_put_price_entry) / abs(strangle_price_entry)) |>
    reframe(
        min_date = min(date_entry),
        max_date = max(date_entry),
        med = median(put_pct_of_strangles),
        mean = mean(put_pct_of_strangles),
        min = min(put_pct_of_strangles),
        max = max(put_pct_of_strangles),
        sd = round(sd(put_pct_of_strangles), 2),
        n = n()
    ) |>
    mutate(across(med:max, function(x) scales::percent(x)))

pct_of_strangle_plot <- pnls_tbl |>
    select(date_entry, strangle_price_entry, tail_put_price_entry) |>
    mutate(put_pct_of_strangles = abs(tail_put_price_entry) / abs(strangle_price_entry)) |>
    ggplot(aes(put_pct_of_strangles, fill = "name")) +
    geom_histogram(bins = 100) +
    scale_x_continuous(labels = scales::percent) +
    labs(
        title = "Put Entry Price Percentage of Strangle Entry Price",
        subtitle = sprintf(
            "(n=%s, µ=%s, min=%s, max=%s, sd=%s)",
            pct_of_strangle_summaries$n,
            pct_of_strangle_summaries$mean,
            pct_of_strangle_summaries$min,
            pct_of_strangle_summaries$max,
            pct_of_strangle_summaries$sd
        ),
        x = "Put Entry Price Percentage of Strangle Entry Price",
        y = "Frequency",
        caption = str_wrap(
            "Put (7dte, 2delta) entry price percentage of strangle (7dte, 20delta) entry price"
        )
    )

pct_of_strangle_plot +
    scale_fill_manual(values = viridis::viridis(1, option = "E")) +
    theme(legend.position = "none")
```

```{r include=FALSE, eval=FALSE}
styled_pct_of_strangle_plot <- pct_of_strangle_plot +
    scale_fill_manual(values = nedhmn_palette) +
    nedhmn_theme()

ggsave(
    filename = "apps/spy-strangles-with-put-hedge/public/put-percentage-of-strangles-distribution.png",
    plot = styled_pct_of_strangle_plot,
    height = 4,
    width = 8
)
```

## Comparison Analysis

### Summary Statistics

```{r}
pnls_tbl |>
    select(date_entry, short_strangle_dollar_pnl, position_dollar_pnl) |>
    pivot_longer(contains("pnl")) |>
    group_by(name) |>
    reframe(
        sum = sum(value),
        min = min(value),
        q1 = quantile(value, 0.25)[1],
        med = median(value),
        mean = mean(value),
        q3 = quantile(value, 0.75)[1],
        max = max(value),
        stddev = sd(value),
        n = n()
    )
```

### Hedged Short Strangle vs Short Strangle 

```{r}
cum_pnls_tbl <- pnls_tbl |>
    select(date_entry, short_strangle_dollar_pnl, position_dollar_pnl) |>
    arrange(date_entry) |>
    mutate(across(contains("pnl"), function(x) cumsum(x), .names = "cum_{.col}"))

cum_pnls_plot <- cum_pnls_tbl |>
    pivot_longer(contains("cum")) |>
    ggplot(aes(x = date_entry, y = value * 100, color = name)) +
    geom_line() +
    scale_y_continuous(
        labels = scales::dollar
    ) +
    labs(
        title = "Short Strangle PnL With and Without Tail Put Hedge",
        subtitle = sprintf("SPY data from %s to %s", min(cum_pnls_tbl$date_entry), max(cum_pnls_tbl$date_entry)),
        x = "",
        y = "Cumulative PnL",
        color = "Legend",
        caption = str_wrap(
            "Short strangles are 20-delta, 7-dte, SPY strangles. Hedged short strangles adds a long 2-delta put. Each leg is 1 lot.",
            width = 100
        )
    )

cum_pnls_plot +
    scale_color_manual(
        values = viridis::viridis(3)[c(1, 2)],
        labels = c(
            "cum_position_dollar_pnl" = "Hedged Short Strangle",
            "cum_short_strangle_dollar_pnl" = "Short Strangle"
        )
    )
```

```{r include=FALSE, eval=FALSE}
styled_cum_pnls_plot <- cum_pnls_plot +
    scale_color_manual(
        values = nedhmn_palette[c(1, 3)],
        labels = c(
            "cum_position_dollar_pnl" = "Hedged Short Strangle",
            "cum_short_strangle_dollar_pnl" = "Short Strangle"
        )
    ) +
    nedhmn_theme(show_legend = TRUE)

ggsave(
    filename = "apps/spy-strangles-with-put-hedge/public/short-strangles-comparison-line-chart.png",
    plot = styled_cum_pnls_plot,
    height = 4,
    width = 8
)
```

### Summary Table

```{r}
pnls_tbl |>
    select(date_entry, short_strangle_dollar_pnl, position_dollar_pnl) |>
    pivot_longer(contains("pnl"), names_to = "strategy") |>
    mutate(value = value * 100) |>
    group_by(strategy) |>
    reframe(
        total = sum(value),
        min_pnl = min(value),
        mean_pnl = mean(value),
        sd = sd(value),
        n = n()
    ) |>
    mutate(
        strategy = ifelse(strategy == "position_dollar_pnl", "Hedged Short Strangle", "Short Strangle"),
        across(total:sd, function(x) scales::dollar(round(x, 2)))
    )
```

## Hedged Short Strangle Strategy 

### Circle Trades When Tail Put Made Money

```{r}
circled_hedged_plot <- cum_pnls_tbl |>
    ggplot(aes(x = date_entry, y = cum_position_dollar_pnl * 100, color = "name")) +
    geom_line() +
    geom_point(
        data = cum_pnls_tbl[cum_pnls_tbl$position_dollar_pnl > cum_pnls_tbl$short_strangle_dollar_pnl, ],
        color = nedhmn_palette[4],
        size = 5,
        stroke = 1.5,
        shape = 1
    ) +
    scale_y_continuous(labels = scales::dollar) +
    labs(
        title = "Short Strangle with Circled Tail Put Hedge",
        subtitle = sprintf(
            "%s to %s",
            min(cum_pnls_tbl$date_entry),
            max(cum_pnls_tbl$date_entry)
        ),
        x = "",
        y = "Cumulative PnL",
        caption = str_wrap(
            "SPY 7dte 20delta short strangles with long 7dte 2delta put hedge. The circles are the times that the put made money.",
            width = 100
        )
    )

circled_hedged_plot +
    scale_color_manual(values = viridis::viridis(1, option = "E")) +
    theme(legend.position = "none")
```

```{r include=FALSE, eval=FALSE}
styled_circled_hedged_plot <- circled_hedged_plot +
    scale_color_manual(values = nedhmn_palette) +
    nedhmn_theme()

ggsave(
    filename = "apps/spy-strangles-with-put-hedge/public/circled-hedged-short-strangle-pnl.png",
    plot = styled_circled_hedged_plot,
    height = 4,
    width = 8
)
```
